---
title: "Data Camp Tree Specie Challenge"
author: "Omotola_Olasope"
date: "`r Sys.Date()`"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
## Contents
 - [Introduction and Background](#introduction-and-background)
  * [Objectives](#objectives)
 - [Data Preparation](#data-preparation)
 - [Data Processing/Cleaning](#data-processingcleaning)
 - [Analysis with R](#analysis-with-r)
 - [Key findings and Summary](#key-findings-and-summary)
  * [Recommendations](#recommendations)
 
## Introduction and background
I work for a nonprofit organization advising the planning department on ways to improve the quantity and quality of trees in New York City. The urban design team believes tree size (using trunk diameter as a proxy for size) and health are the most desirable characteristics of city trees.

The city would like to learn more about which tree species are the best choice to plant on the streets of Manhattan.

### Objectives
The aim of the analysis is to create a report that covers the following:

- What are the most common tree species in Manhattan?
- Which are the neighborhoods with the most trees?
- A visualization of Manhattan's neighborhoods and tree locations.
- What ten tree species would you recommend the city plant in the future?

## Data Preparation
The Tree census and neighborhood information is publicly available from the City of New York NYC Open Data <https://opendata.cityofnewyork.us/data/>

## Data Processing/Cleaning

```{r include=FALSE, results='hide', message=FALSE, warning=FALSE}

## Installing and loading common packages and libraries
library(tidyverse)
library(dplyr)

treesdf <- read.csv('trees.csv')

## Exploring the data
head(treesdf)
colnames(treesdf)
print(treesdf)

```

The data was relatively clean. However, it was observed that all the dead trees had no common name for its species, so I separated the dead trees from the alive ones to get more detailed results based on the projects objectives.

```{r include=FALSE, results='hide', message=FALSE, warning=FALSE}

#seperate dead trees from alive ones
dead_trees <- treesdf %>% 
  filter(status == "Dead")

alive_trees <- treesdf %>% 
  filter(status == "Alive")

```

## Analysis With R

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}

# Get number of alive trees per specie
tree_count <- alive_trees %>% 
  count(spc_common)

# change column name for n column
colnames(tree_count)[2]  <- "number_of_trees"

# Sort by number_of_trees 
tree_count <- tree_count[order(tree_count$number_of_trees, decreasing=TRUE, na.last=FALSE),]

# Most populous trees
tree_count <- tree_count %>% 
  arrange(-number_of_trees)

head(tree_count)

```

```{r}

# Get number of alive trees per location
nta_location_count <- alive_trees %>% 
  count(nta_name)

# change column name for n column
colnames(nta_location_count)[2]  <- "number_of_trees"

# Sort by number_of_trees 
nta_location_count <- nta_location_count[order(nta_location_count$number_of_trees, decreasing=TRUE, na.last=FALSE),]

# View location with most populous trees
nta_location_count <- nta_location_count %>% 
  arrange(-number_of_trees)

head(nta_location_count)
tail(nta_location_count)

```

```{r}

# Get number of dead trees per location
deadTrees.location_count <- dead_trees %>% 
  count(nta_name)

# change column name for n column
colnames(deadTrees.location_count)[2]  <- "number_of_trees"

# Sort by number_of_trees 
deadTrees.location_count <- deadTrees.location_count[order(deadTrees.location_count$number_of_trees, decreasing=TRUE, na.last=FALSE),]

# View location with most populous dead trees
deadTrees.location_count <- deadTrees.location_count %>% 
  arrange(-number_of_trees)

head(deadTrees.location_count)

```


Since the urban design team believes trunk diameter and health are the most desirable characteristics of city trees, lets find out the trees with good health status and diameter greater than 9 to determine the best set of trees to recommend.

```{r}

# sort trees by health status
trees_by_health <- alive_trees %>% 
  group_by(spc_common, health) %>% 
  summarise(number = n())

trees_by_healthWIDE <- trees_by_health %>% 
  spread(health, number)

## add row sums to get percentages
trees_by_healthWIDE$row_sum <- rowSums(trees_by_healthWIDE[ , c(2,3,4)], na.rm=TRUE)

trees_by_healthWIDE$percentage_good <- round(100*(trees_by_healthWIDE$Good/trees_by_healthWIDE$row_sum), 2)

trees_by_healthWIDE$percentage_fair <- round(100*(trees_by_healthWIDE$Fair/trees_by_healthWIDE$row_sum), 2)

trees_by_healthWIDE$percentage_poor <- round(100*(trees_by_healthWIDE$Poor/trees_by_healthWIDE$row_sum), 2)

# sort trees by diameter
tree_diameter <- alive_trees %>% 
  group_by(spc_common) %>% 
  summarise(
    mean_diameter = mean(tree_dbh),
    median_diameter = median(tree_dbh),
    number = n()) %>%
  arrange(-mean_diameter)

# merge the two sorted data frames by specie
tree_status.diameter_health <- merge(tree_diameter, trees_by_healthWIDE, by="spc_common")

# clean merged dataframe
## remove column
tree_status.diameter_health <- tree_status.diameter_health[,-4]
tree_status.diameter_health <- tree_status.diameter_health[,-3]

## change column name
colnames(tree_status.diameter_health)[3]  <- "fair_health"
colnames(tree_status.diameter_health)[4]  <- "good_health"
colnames(tree_status.diameter_health)[5]  <- "poor_health"
colnames(tree_status.diameter_health)[6]  <- "total_number_of_trees"


#filter by diameter and percentage good to determine the best trees to recommend
recommended_trees <- tree_status.diameter_health %>% 
  filter(mean_diameter >= 9 & percentage_good >= 70) %>% 
  arrange(-percentage_good)

# View first 10 recommended trees to plant
head(recommended_trees, 10)

```

# Key Findings and Summary
A visualization of Manhattan's neighborhoods and tree locations can be found here. <https://public.tableau.com/app/profile/omotola.olasope/viz/AvisualizationofManhattansneighborhoodsandtreelocations_/Map?publish=yes>

### Table One
```{r echo=FALSE, options(knitr.duplicate.label = "allow")}

head(tree_count)

```

### Description
The most common tree specie in Manhattan is honeylocust. Others include Callery pear, ginkgo, pin oak and Sophora.

### Table Two
```{r echo=FALSE}

head(nta_location_count)

```

### Description
The neighborhood with the most trees is Upper West Side. Other neighborhoods include Upper East Side-Carnegie Hill, West Village, Central Harlem North-Polo Grounds and Hudson Yards-Chelsea-Flatiron-Union Square.

### Table Three
```{r echo=FALSE}

head(recommended_trees, 10)

```

### Description
Ten tree species I would recommend the city should plant in the future based on their trunk diameter and health are Smoketree, Black Maple, Amur Tork Tree, Willow Oak, Siberian Elm, Honeylocust, Pin Oak, American Elm, White Ash and Sophora. 

## Recommendations
* The city should plant more of the following tree species: Smoketree, Black Maple, Amur Tork Tree, Willow Oak, Siberian Elm, Honeylocust, Pin Oak, American Elm, White Ash and Sophora.

* More trees need to be planted in the following locations Stuyvesant Town-Cooper Village, Manhattanville, Midtown-Midtown South, Gramercy, Battery Park City-Lower Manhattan and Chinatown.

* Further analysis should be carried out to ascertain the reason for the number of dead trees in the following regions Washington Heights South, Hudson Yards-Chelsea-Flatiron-Union Square, Central Harlem North-Polo Grounds, East Harlem North, Washington Heights North, and West Village.

------
